\textit{k}-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean

### 2023-07-20 additional_files/manuscript.tex
Final files sent for publication

\section*{Introduction}
Genome-wide association studies (GWAS) are analytical approaches that detect
statistical associations between phenotypic observations for a trait of
interest and the genotypes of variants distributed throughout the
genome \shortcite{tam2019}. This approach has been used for over 15 years
\shortcite{visscher2012} and has enabled significant advances in our understanding
of the genetic architecture of traits involved in human health
\shortcite<e.g.>[]{wellcome2007gwas} and crop \shortcite{tibbs2021} and animal
\shortcite{ma2019} production, among others.

@@

\section*{Materials and methods}

\subsection*{Sample selection and processing of sequencing data}

@@

\subsection*{Discovery and genotyping of SVs}

@@

\subsection*{Discovery and genotyping of SNPs and indels}

@@

\subsection*{Computing the presence/absence table of \emph{k}-mers}

@@

\subsection*{Phenotypic data}

@@

\subsection*{GWAS analyses}

@@

\subsection*{Analysis of significant \emph{k}-mers}

@@

\subsection*{Downstream analyses of significant association signals}

@@

\subsection*{Linking significant \emph{k}-mers to variants}

@@

\subsection*{Computation of LD between \emph{k}-mers}

@@

\subsection*{Software used}
\label{sv-gwas-software-used}

@@

\section*{Results and discussion}

@@

\subsection*{Loci with known genes or causal variants}

We first wanted to assess the performance of the various GWAS
approaches by analyzing the results at loci for which the genes have
already been cloned. These loci provide interesting test cases because
the expected results (causal genes or variants) are known. Table \ref{loci-table} shows
a summary of the results discussed below.

\subsubsection*{Flower color -- \textit{W1} locus}

\shortciteA{zabala2007} identified the flavonoid 3’5’-hydroxylase (F3’5’H) gene
Glyma.13g\-072100 as associated with the \textit{W1} locus for flower color. In
Williams82 and other accessions with white flowers, a 65-bp insertion into the
third exon of the gene results in a premature stop codon that renders the
F3’5’H enzyme non-functional and prevents pigmentation of the flower. All
GWAS approaches detected a signal overlapping the gene at this locus (Supplemental Figures
S2 and S3). The most
significant SV and \textit{k}-mers corresponded to the known causal variant
(Figure \ref{flower-color-main-figure}a-b and Supplemental Figure S4c-d).
In particular, the assembly of the reads containing the most significant \textit{k}-mers
and the alignment of the resulting haplotypes showed that these \textit{k}-mers
directly tagged the causal variant (Figure \ref{flower-color-main-figure}c-e).
As the catalogue of variants analyzed by Platypus contains only SNPs and small
indels, the causal variant could not be captured by this approach. Still, SNPs
and indels located within the gene sequence were among the most significantly
associated ones (Supplemental Figure S4b).

% Figure 1
\begin{figure}
	\centering
	\includegraphics[width = 5in]{flower_color_W1_main_figure.png}
	\caption{(Caption on the following page.)}
	\label{flower-color-main-figure}
\end{figure}

\begin{figure}
	\contcaption{Results of SV- and \textit{k}-mer-based GWAS at the
	\textit{W1} locus for flower color. (a) Genome-wide Manhattan plots.
	Horizontal dotted lines indicate the significance
	threshold whereas vertical dotted lines indicate the location of the
	\textit{W1} locus. (b) Zoomed-in Manhattan plots at the \textit{W1} locus.
	The vertical dotted lines indicate the location of the
	Glyma.13g072100 gene associated with this locus. The shaded gray
	rectangles show the regions delimited by the top 5\% (SVs) and
	1\% (\textit{k}-mers) associations.  (c) Gene model of Glyma.13g072100.
	Rectangles represent exons, colored rectangles represent coding
	sequences, and arrows indicate the direction of transcription. The red
	rectangle highlights the region shown in panel (d). (d) Pairwise
	alignment of the two haplotype sequences in the population at the
	\textit{W1} locus. Nucleotides are
	colored according to the \textit{p}-value of the most significant
	\textit{k}-mer overlapping them. Vertical dotted lines between haplotypes
	indicate sequence differences, whereas dashes indicate gaps in the
	alignment. (e) Contingency table of the phenotypes and haplotypes found
	in the population at the \textit{W1} locus.}
\end{figure}

@@

\subsubsection*{Pubescence color -- \textit{T} locus}

@@

\subsubsection*{Pubescence color -- \textit{Td} locus}

% Pubescence color - Td locus
We conducted a separate GWAS analysis contrasting only accessions with tawny
and light tawny pubescence because the \textit{Td} locus is known to control
these differences in color. Accordingly, we detected signals in this region
using all three approaches, but only SNP- and \textit{k}-mer-based GWAS
identified a signal that overlapped Glyma.03g258700 (Figure
\ref{pubescence-color-main-figure}a-b, and Supplemental Figures S10 and
S11).  While the most significant
\textit{k}-mer identified the causal variant documented at this locus
\shortcite{yan2020}, it was the fourth most significantly associated variant for
the SNP-based approach (Figure \ref{pubescence-color-main-figure}b-d, Supplemental Figure 
S12).  Interestingly, some of the accessions
bearing the haplotype associated with tawny pubescence displayed light tawny
pubescence, suggesting that there may be more than one causal variant at this
locus (Figure \ref{pubescence-color-main-figure}e).

% Figure 2
\begin{figure}
	\centering
	\includegraphics[width = 5in]{pubescence_color_nogray_Td_main_figure.png}
	\caption{(Caption on the following page.)}
	\label{pubescence-color-main-figure}
\end{figure}

\begin{figure}
	\contcaption{Results of SNP- and \textit{k}-mer-based GWAS at the
	\textit{Td} locus for pubescence color. (a) Genome-wide Manhattan plots.
	Horizontal dotted lines indicate the significance
	threshold whereas vertical dotted lines indicate the location of the
	\textit{Td} locus. (b) Zoomed-in Manhattan plots at the \textit{Td} locus.
	The vertical dotted lines indicate the location of the
	Glyma.03g258700 gene associated with this locus. The shaded gray
	rectangles show the regions delimited by the top 1\% associations.
	In the case of SNPs, blue points denote markers used in the
	original analysis, whereas red points denote markers that had originally
	been pruned but whose \emph{p}-values were computed after signal
	discovery.
	(c) Gene model of Glyma.03g258700. Rectangles represent exons, colored
	rectangles represent coding sequences, and the arrow indicates the
	direction of transcription. The red rectangle highlights the region
	shown in panel (d). (d) Pairwise alignment of the two haplotype
	sequences in the population at the \textit{Td} locus. Nucleotides are
	colored according to the
	\textit{p}-value of the most significant \textit{k}-mer overlapping
	them. The vertical dotted line between haplotypes indicates the location of
	the causal SNP at this locus. (e) Contingency table of
	the phenotypes and haplotypes found in the population at the
	\textit{Td} locus.}
\end{figure}

\subsubsection*{Seed coat color - \textit{I} locus}

@@

\subsubsection*{Seed coat color - \textit{G} locus}

We conducted a second GWAS analysis contrasting accessions with yellow and
green seed coats in order to target the \textit{G} locus. Accordingly, this
analysis found strong signals overlapping the \emph{G} locus using all three
approaches (Supplemental Figure S15).
\shortciteA{wang2018} identified Glyma.01g198500 as the gene associated with
this locus and found an A>G SNP affecting transcript splicing as the causal
variant. Consistently with their results, we identified the two
most significant \emph{k}-mers and the second most significant marker from the
SNP-based analysis as corresponding to that causal variant (Figure
\ref{seed-coat-color-main-figure} and Supplemental Figures
S16 and
S17). Unsurprisingly, SV-based GWAS
did not identify the causal variant given its type.

% Figure 3 with the results on pubescence color at locus G
\begin{figure}
	\centering
	\includegraphics[width = 5in]{seed_coat_color_greenyellow_G_main_figure.png}
	\caption{(Caption on the following page.)}
	\label{seed-coat-color-main-figure}
\end{figure}

\begin{figure}
	\contcaption{Results of SNP- and \textit{k}-mer-based GWAS at the
	\textit{G} locus for seed coat color. (a) Genome-wide Manhattan plots.
	Horizontal dotted lines indicate the significance
	threshold whereas vertical dotted lines indicate the location of the
	\textit{G} locus. (b) Zoomed-in Manhattan plots at the \textit{G} locus.
	The vertical dotted lines indicate the location of the
	Glyma.01g198500 gene associated with this locus. The shaded gray
	rectangles show the regions delimited by the top 1\% associations.
	In the case of SNPs, blue points denote markers used in the
	original analysis, whereas red points denote markers that had originally
	been pruned but whose \emph{p}-values were computed after signal
	discovery.
	(c) Gene model of Glyma.01g198500. Rectangles represent exons, colored
	rectangles represent coding sequences, and the arrows indicate the
	direction of transcription. The red rectangle highlights the region
	shown in panel (d). (d) Pairwise alignment of the two haplotype
	sequences in the population at the \textit{G} locus. Nucleotides are
	colored according to the
	\textit{p}-value of the most significant \textit{k}-mer overlapping
	them. The vertical dotted line between haplotypes indicates the location of
	the causal SNP at this locus. (e) Contingency table of
	the phenotypes and haplotypes found in the population at the
	\textit{G} locus.}
\end{figure}

\subsubsection*{Stem termination type - \textit{Dt1} locus}

@@

\subsubsection*{Hilum color -- \textit{R} locus}

@@

\subsubsection*{Pubescence density -- \textit{Ps} locus}

We conducted a GWAS analysis comparing accessions with normal and semi-sparse
pubes\-cence in order to detect loci associated with pubescence density. This analysis
detected signals at the \textit{Ps} locus using all three approaches (Figure
\ref{pubescence-density-main-figure}a-b and Supplemental Figures S26, S27
and S28). The causal
variant at this recently cloned locus is a copy number variant (CNV) overlapping the
Glyma.12g187200 gene, with higher copy number resulting in reduced pubescence \shortcite{liu2020ps}.
We identified a region of $\sim$42 kb that varied in copy number across
the population and was associated with the trait (Figure
\ref{pubescence-density-main-figure}c), and therefore possibly corresponded to the
25.6-kb region identified as the causal CNV by \shortciteA{liu2020ps}. We found that
reads containing the most significantly associated \textit{k}-mer mapped to either of the two ends
of the CNV (Figure \ref{pubescence-density-main-figure}d). This suggests that
the most associated \textit{k}-mer is indeed linked to the CNV
that is causal at this locus, while \textit{k}-mers within the
CNV region may represent differences in sequence between the copies.

% Figure 4 with the results on pubescence density at locus Ps
\begin{figure}
	\centering
	\includegraphics[width = 6in]{pubescence_density_Ps_main_figure.png}
	\caption[Results of \textit{k}-mer-based GWAS at the
	\textit{Ps} locus for pubescence density.]{Results of \textit{k}-mer-based GWAS at the
	\textit{Ps} locus for pubescence density.
	(a) Genome-wide Manhattan plot showing the location of significant \textit{k}-mers. The
	horizontal dotted line indicates the significance threshold whereas the
	vertical dotted line indicates the location of the \textit{Ps} locus.
	(b) Zoomed-in Manhattan plot at the \textit{Ps} locus. The
	vertical dotted line indicates the location of the most significant
	\textit{k}-mer. The shaded gray area shows the location of the causal CNV.
	(c) Estimated copy number (sequencing depth at position divided by
	accession average sequencing depth) of accessions with normal and
	semi-sparse pubescence at the \textit{Ps} locus. Solid lines represent
	the median copy number across 73 (semi-sparse pubescence) and 180
	accessions (normal pubescence) computed using a 100-bp sliding window.
	Shaded areas outline the first and third quartiles. The start and end
	of the CNV region can be clearly inferred from the divergence between
	the copy number of accessions with contrasting pubescence.
	(d) Histogram showing the location of 1,169 mapped reads containing the
	most significant \textit{k}-mer in accessions with semi-sparse pubescence.
	Reads containing this \textit{k}-mer map to either end of the CNV
	region, suggesting that they are associated with this variant. Panels
	(b), (c) and (d) share the same x-axis.}
	\label{pubescence-density-main-figure}
\end{figure}

@@

\subsubsection*{Seed coat luster -- \textit{B} locus}

@@

\subsection*{Analysis of loci with unknown causal genes}

@@

\subsubsection*{Pod color -- \textit{L1} and \textit{L2} loci}

@@

\subsubsection*{Pubescence form -- \textit{Pa1} and \textit{Pa2} loci}
\label{sv-gwas-main-results-pubescence-form-pa1-pa2}

The two known loci for pubescence form in soybean, \textit{Pa1} and
\textit{Pa2}, have also not been cloned yet. Nevertheless, \shortciteA{gilbert2017}
has suggested Glyma.12g213900 as a candidate for \textit{Pa1} based on data
published by \shortciteA{bandillo2017}. This gene encodes a MYB transcription factor
annotated as playing a role in trichome branching. Indeed, the most significant
signals found by all three approaches were located near this gene (Figure
\ref{pubescence-form-main-figure}a-b, and Supplemental Figures S35 and
S36).  The analyses based on SNPs and
\emph{k}-mers identified two highly significant non-synonymous SNPs in
Glyma.12g213900 that were associated with appressed pubescence (Figure
\ref{pubescence-form-main-figure}c-d and Supplemental Figures S37 and
S38). The alternative alleles for these two
SNPs co-occurred together in our dataset and may represent causal variants at
the \emph{Pa1} locus. The signals observed at the \emph{Pa2} locus were much
weaker and the signal observed from the \emph{k}-mer GWAS overlapped a total of
20 genes (Supplemental Figures S35 and
S39), none of which appeared obviously linked
to pubescence form.

% Figure 5 with the results on pubescence form at locus Pa1
\begin{figure}
	\centering
	\includegraphics[width = 5in]{pubescence_form_all_Pa1_main_figure.png}
	\caption{(Caption on the following page.)}
	\label{pubescence-form-main-figure}
\end{figure}

\begin{figure}
	\contcaption{Results of SNP- and \textit{k}-mer-based GWAS at the
	\textit{Pa1} locus for pubescence form. (a) Genome-wide Manhattan plots.
	Horizontal dotted lines indicate the significance
	threshold whereas vertical dotted lines indicate the location of the
	\textit{Pa1} and \textit{Pa2} loci. (b) Zoomed-in Manhattan plots at the \textit{Pa1} locus.
	The vertical dotted lines indicate the location of the
	Glyma.12g213900 gene suggested as a candidate for this locus. The shaded gray
	rectangles show the regions delimited by the top 1\% associations.
	In the case of SNPs, blue points denote markers used in the
	original analysis, whereas red points denote markers that had originally
	been pruned but whose \emph{p}-values were computed after signal
	discovery.
	(c) Gene model of Glyma.12g213900. Rectangles represent exons, colored
	rectangles represent coding sequences, and the arrows indicate the
	direction of transcription.
	(d) Manhattan plots at the location of the Glyma.12g213900 gene.
	Vertical lines indicate the location of two non-synonymous
	SNPs associated with appressed pubescence, which may represent causal variants
	at this locus. Panels (c) and (d) share the same x-axis.}
\end{figure}

@@

\subsubsection*{Resistance to \textit{P. sojae}}

@@

\subsubsection*{Novel loci for stem termination and pod color}

In addition to verifying known loci, we also observed potentially novel loci in
the \textit{k}-mer GWAS analysis of stem termination and pod color.  When
comparing accessions with indeterminate and semi-determinate stem termination
types, we observed previously undocumented signals on chromosomes Gm11, Gm16
and Gm18 in addition to the known \textit{Dt1} locus (Figure
\ref{stem-termination-main-figure}a).  Analysis of the pairwise LD between
significant \textit{k}-mers found no LD between these signals, suggesting that
they may represent \textit{bona fide} loci associated with stem termination
type (Figure \ref{stem-termination-main-figure}b). We note that the signal newly
identified on Gm18 is located 5 Mb away from the already documented \textit{Dt2}
locus and may therefore correspond to that locus.
Similarly, a \textit{k}-mer
GWAS analysis comparing accessions with black and brown pods found a previously
undocumented signal on chromosome Gm15 (Supplemental Figure S43).
\textit{k}-mers at this new locus
showed no LD with \textit{k}-mers at the known \textit{L1} locus and may
therefore represent a distinct locus (Supplemental Figure 
S44). The observed signal overlaps a total of 27 genes
(Supplemental Figure S45), but none of those are annotated
as having an obvious role in controlling pod color.

% Figure 6 with the results novel loci for stem termination type
\begin{figure}
	\centering
	\includegraphics[width = 6in]{stem_termination_sn_main_figure.png}
	\caption[Detection of novel signals for stem termination type
	using a \textit{k}-mer GWAS analysis contrasting accessions with
	indeterminate and semi-determinate stems]{Detection of novel signals for stem termination type
	using a \textit{k}-mer-based GWAS analysis contrasting accessions with
	indeterminate and semi-determinate stems.
	(a) Genome-wide Manhattan plot showing the location of significant \textit{k}-mers.
	The horizontal dotted line indicates the significance threshold
	whereas vertical dotted lines indicate the location of the \textit{Dt1}
	classical locus as well as three new loci of interest (\textit{stGm11},
	\textit{stGm16}, and \textit{stGm18}).
	(b) Pairwise LD between the 162 significant \textit{k}-mers identified.
	\textit{k}-mers are sorted along the axes according to their
	putative position on the reference assembly version 4 of Williams82.
	\textit{k}-mers are
	represented in the same order along the x- and y-axis. The blue
	rectangles below the x-axis represent the
	-log\textsubscript{10}(\textit{p}) of each \textit{k}-mer.}
	\label{stem-termination-main-figure}
\end{figure}

\subsection*{Potential of \emph{k}-mer-based GWAS}

One of the most striking results from the analyses presented above is the
impressive performance of \emph{k}-mer-based GWAS compared to SNP- and SV-based
GWAS. The GWAS analyses using \emph{k}-mers identified the causal variant as
the most significantly associated variant at four loci with various types of
causal variants: \emph{W1} for flower color, \emph{Td} for pubescence color,
\emph{G} for seed coat color, and \emph{Ps} for pubescence density (Table
\ref{loci-table}). In comparison, this result was achieved by SV-based GWAS at
the \textit{W1} locus only, and at none of the loci for SNP-based GWAS.
The \textit{k}-mer approach also systematically performed at least as well as
the other two approaches in identifying signals overlapping known variants or
genes (Table \ref{loci-table}). Moreover, whenever \textit{k}-mer-based GWAS
did detect a signal, the region defined by the 1\% most significant
\textit{k}-mers overlapped the known causal variant or gene.

@@

\subsection*{Limitations and challenges of \emph{k}-mer-based GWAS}

@@

\subsection*{Use of SV genotypes for GWAS}

@@

\section*{Conclusions}

@@

\begin{description}
	\item[Supplemental file 1.pdf] Supplemental methods and results, including
		Supplemental Tables S1 to S11 and Supplemental Figures S1 to
		S60.

@@

>>>>

\section{Introduction}
Genome-wide association studies (GWAS) are analytical approaches that detect
statistical associations between phenotypic observations for a trait of
interest and the genotypes of variants distributed throughout the
genome \shortcite{tam2019}. This approach has been used for over 15 years
\shortcite{visscher2012} and has enabled significant advances in our understanding
of the genetic architecture of traits involved in human health
\shortcite<e.g.>[]{wellcome2007gwas} and crop \shortcite{tibbs2021} and animal
\shortcite{ma2019} production, among others.

@@

\section{Materials and methods}

\subsection{Sample selection and processing of sequencing data}

@@

\subsection{Discovery and genotyping of SVs}

@@

\subsection{Discovery and genotyping of SNPs and indels}

@@

\subsection{Computing the presence/absence table of \emph{k}-mers}

@@

\subsection{Phenotypic data}

@@

\subsection{GWAS analyses}

@@

\subsection{Analysis of significant \emph{k}-mers}

@@

\subsection{Downstream analyses of significant association signals}

@@

\subsection{Linking significant \emph{k}-mers to variants}

@@

\subsection{Computation of LD between \emph{k}-mers}

@@

\subsection{Software used}
\label{sv-gwas-software-used}

@@

\section{Results and discussion}

@@

\subsection{Loci with known genes or causal variants}

We first wanted to assess the performance of the various GWAS
approaches by analyzing the results at loci for which the genes have
already been cloned. These loci provide interesting test cases because
the expected results (causal genes or variants) are known. Table 1 shows
a summary of the results discussed below.

\subsubsection{Flower color -- \textit{W1} locus}

\shortciteA{zabala2007} identified the flavonoid 3’5’-hydroxylase (F3’5’H) gene
Glyma.13g\-072100 as associated with the \textit{W1} locus for flower color. In
Williams82 and other accessions with white flowers, a 65-bp insertion into the
third exon of the gene results in a premature stop codon that renders the
F3’5’H enzyme non-functional and prevents pigmentation of the flower. All
GWAS approaches detected a signal overlapping the gene at this locus (Supplemental Figures
S2 and S3). The most
significant SV and \textit{k}-mers corresponded to the known causal variant
(Figure 1a-b and Supplemental Figure S4c-d).
In particular, the assembly of the reads containing the most significant \textit{k}-mers
and the alignment of the resulting haplotypes showed that these \textit{k}-mers
directly tagged the causal variant (Figure 1c-e).
As the catalogue of variants analyzed by Platypus contains only SNPs and small
indels, the causal variant could not be captured by this approach. Still, SNPs
and indels located within the gene sequence were among the most significantly
associated ones (Supplemental Figure S4b).

@@

\subsubsection{Pubescence color -- \textit{T} locus}

@@

\subsubsection{Pubescence color -- \textit{Td} locus}

% Pubescence color - Td locus
We conducted a separate GWAS analysis contrasting only accessions with tawny
and light tawny pubescence because the \textit{Td} locus is known to control
these differences in color. Accordingly, we detected signals in this region
using all three approaches, but only SNP- and \textit{k}-mer-based GWAS
identified a signal that overlapped Glyma.03g258700 (Figure
2a-b, and Supplemental Figures S10 and
S11).  While the most significant
\textit{k}-mer identified the causal variant documented at this locus
\shortcite{yan2020}, it was the fourth most significantly associated variant for
the SNP-based approach (Figure 2b-d, Supplemental Figure 
S12).  Interestingly, some of the accessions
bearing the haplotype associated with tawny pubescence displayed light tawny
pubescence, suggesting that there may be more than one causal variant at this
locus (Figure 2e).

\subsubsection{Seed coat color - \textit{I} locus}

@@

\subsubsection{Seed coat color - \textit{G} locus}

We conducted a second GWAS analysis contrasting accessions with yellow and
green seed coats in order to target the \textit{G} locus. Accordingly, this
analysis found strong signals overlapping the \emph{G} locus using all three
approaches (Supplemental Figure S15).
\shortciteA{wang2018} identified Glyma.01g198500 as the gene associated with
this locus and found an A>G SNP affecting transcript splicing as the causal
variant. Consistently with their results, we identified the two
most significant \emph{k}-mers and the second most significant marker from the
SNP-based analysis as corresponding to that causal variant (Figure
3 and Supplemental Figures
S16 and
S17). Unsurprisingly, SV-based GWAS
did not identify the causal variant given its type.

\subsubsection{Stem termination type - \textit{Dt1} locus}

@@

\subsubsection{Hilum color -- \textit{R} locus}

@@

\subsubsection{Pubescence density -- \textit{Ps} locus}

We conducted a GWAS analysis comparing accessions with normal and semi-sparse
pubes\-cence in order to detect loci associated with pubescence density. This analysis
detected signals at the \textit{Ps} locus using all three approaches (Figure
4a-b and Supplemental Figures S26, S27
and S28). The causal
variant at this recently cloned locus is a copy number variant (CNV) overlapping the
Glyma.12g187200 gene, with higher copy number resulting in reduced pubescence \shortcite{liu2020ps}.
We identified a region of $\sim$42 kb that varied in copy number across
the population and was associated with the trait (Figure
4c), and therefore possibly corresponded to the
25.6-kb region identified as the causal CNV by \shortciteA{liu2020ps}. We found that
reads containing the most significantly associated \textit{k}-mer mapped to either of the two ends
of the CNV (Figure 4d). This suggests that
the most associated \textit{k}-mer is indeed linked to the CNV
that is causal at this locus, while \textit{k}-mers within the
CNV region may represent differences in sequence between the copies.

@@

\subsubsection{Seed coat luster -- \textit{B} locus}

@@

\subsection{Analysis of loci with unknown causal genes}

@@

\subsubsection{Pod color -- \textit{L1} and \textit{L2} loci}

@@

\subsubsection{Pubescence form -- \textit{Pa1} and \textit{Pa2} loci}
\label{sv-gwas-main-results-pubescence-form-pa1-pa2}

The two known loci for pubescence form in soybean, \textit{Pa1} and
\textit{Pa2}, have also not been cloned yet. Nevertheless, \shortciteA{gilbert2017}
has suggested Glyma.12g213900 as a candidate for \textit{Pa1} based on data
published by \shortciteA{bandillo2017}. This gene encodes a MYB transcription factor
annotated as playing a role in trichome branching. Indeed, the most significant
signals found by all three approaches were located near this gene (Figure
5a-b, and Supplemental Figures S35 and
S36).  The analyses based on SNPs and
\emph{k}-mers identified two highly significant non-synonymous SNPs in
Glyma.12g213900 that were associated with appressed pubescence (Figure
5c-d and Supplemental Figures S37 and
S38). The alternative alleles for these two
SNPs co-occurred together in our dataset and may represent causal variants at
the \emph{Pa1} locus. The signals observed at the \emph{Pa2} locus were much
weaker and the signal observed from the \emph{k}-mer GWAS overlapped a total of
20 genes (Supplemental Figures S35 and
S39), none of which appeared obviously linked
to pubescence form.

@@

\subsubsection{Resistance to \textit{P. sojae}}

@@

\subsubsection{Novel loci for stem termination and pod color}

In addition to verifying known loci, we also observed potentially novel loci in
the \textit{k}-mer GWAS analysis of stem termination and pod color.  When
comparing accessions with indeterminate and semi-determinate stem termination
types, we observed previously undocumented signals on chromosomes Gm11, Gm16
and Gm18 in addition to the known \textit{Dt1} locus (Figure
6a).  Analysis of the pairwise LD between
significant \textit{k}-mers found no LD between these signals, suggesting that
they may represent \textit{bona fide} loci associated with stem termination
type (Figure 6b). We note that the signal newly
identified on Gm18 is located 5 Mb away from the already documented \textit{Dt2}
locus and may therefore correspond to that locus.
Similarly, a \textit{k}-mer
GWAS analysis comparing accessions with black and brown pods found a previously
undocumented signal on chromosome Gm15 (Supplemental Figure S43).
\textit{k}-mers at this new locus
showed no LD with \textit{k}-mers at the known \textit{L1} locus and may
therefore represent a distinct locus (Supplemental Figure 
S44). The observed signal overlaps a total of 27 genes
(Supplemental Figure S45), but none of those are annotated
as having an obvious role in controlling pod color.

\subsection{Potential of \emph{k}-mer-based GWAS}

One of the most striking results from the analyses presented above is the
impressive performance of \emph{k}-mer-based GWAS compared to SNP- and SV-based
GWAS. The GWAS analyses using \emph{k}-mers identified the causal variant as
the most significantly associated variant at four loci with various types of
causal variants: \emph{W1} for flower color, \emph{Td} for pubescence color,
\emph{G} for seed coat color, and \emph{Ps} for pubescence density (Table
1). In comparison, this result was achieved by SV-based GWAS at
the \textit{W1} locus only, and at none of the loci for SNP-based GWAS.
The \textit{k}-mer approach also systematically performed at least as well as
the other two approaches in identifying signals overlapping known variants or
genes (Table 1). Moreover, whenever \textit{k}-mer-based GWAS
did detect a signal, the region defined by the 1\% most significant
\textit{k}-mers overlapped the known causal variant or gene.

@@

\subsection{Limitations and challenges of \emph{k}-mer-based GWAS}

@@

\subsection{Use of SV genotypes for GWAS}

@@

\section{Conclusions}

@@

\begin{description}
	\item[Supplemental file 1.pdf] Supplemental methods and results, including
		Supplemental Tables S1 to S12 and Supplemental Figures S1 to
		S61.

@@

\section*{Figure captions}

% Figure 1
Figure 1: Results of SV- and \textit{k}-mer-based GWAS at the \textit{W1} locus
for flower color. (a) Genome-wide Manhattan plots.  Horizontal dotted lines
indicate the significance threshold whereas vertical dotted lines indicate the
location of the \textit{W1} locus.  (b) Zoomed-in Manhattan plots at the
\textit{W1} locus.  The vertical dotted lines indicate the location of the
Glyma.13g072100 gene associated with this locus. The shaded gray rectangles
show the regions delimited by the top 5\% (SVs) and 1\% (\textit{k}-mers)
associations.  (c) Gene model of Glyma.13g072100.  Rectangles represent exons,
colored rectangles represent coding sequences, and arrows indicate the
direction of transcription. The red rectangle highlights the region shown in
panel (d). (d) Pairwise alignment of the two haplotype sequences in the
population at the \textit{W1} locus. Nucleotides are colored according to the
\textit{p}-value of the most significant \textit{k}-mer overlapping them.
Vertical dotted lines between haplotypes indicate sequence differences, whereas
dashes indicate gaps in the alignment. (e) Contingency table of the phenotypes
and haplotypes found in the population at the \textit{W1} locus.

% Figure 2
Figure 2: Results of SNP- and \textit{k}-mer-based GWAS at the \textit{Td}
locus for pubescence color. (a) Genome-wide Manhattan plots.  Horizontal dotted
lines indicate the significance threshold whereas vertical dotted lines
indicate the location of the \textit{Td} locus. (b) Zoomed-in Manhattan plots
at the \textit{Td} locus.  The vertical dotted lines indicate the location of
the Glyma.03g258700 gene associated with this locus. The shaded gray rectangles
show the regions delimited by the top 1\% associations.  In the case of SNPs,
blue points denote markers used in the original analysis, whereas red points
denote markers that had originally been pruned but whose \emph{p}-values were
computed after signal discovery.  (c) Gene model of Glyma.03g258700. Rectangles
represent exons, colored rectangles represent coding sequences, and the arrow
indicates the direction of transcription. The red rectangle highlights the
region shown in panel (d). (d) Pairwise alignment of the two haplotype
sequences in the population at the \textit{Td} locus. Nucleotides are colored
according to the \textit{p}-value of the most significant \textit{k}-mer
overlapping them. The vertical dotted line between haplotypes indicates the
location of the causal SNP at this locus. (e) Contingency table of the
phenotypes and haplotypes found in the population at the \textit{Td} locus.

% Figure 3 with the results on pubescence color at locus G
Figure 3: Results of SNP- and \textit{k}-mer-based GWAS at the \textit{G} locus
for seed coat color. (a) Genome-wide Manhattan plots.  Horizontal dotted lines
indicate the significance threshold whereas vertical dotted lines indicate the
location of the \textit{G} locus. (b) Zoomed-in Manhattan plots at the
\textit{G} locus.  The vertical dotted lines indicate the location of the
Glyma.01g198500 gene associated with this locus. The shaded gray rectangles
show the regions delimited by the top 1\% associations.  In the case of SNPs,
blue points denote markers used in the original analysis, whereas red points
denote markers that had originally been pruned but whose \emph{p}-values were
computed after signal discovery.  (c) Gene model of Glyma.01g198500. Rectangles
represent exons, colored rectangles represent coding sequences, and the arrows
indicate the direction of transcription. The red rectangle highlights the
region shown in panel (d). (d) Pairwise alignment of the two haplotype
sequences in the population at the \textit{G} locus. Nucleotides are colored
according to the \textit{p}-value of the most significant \textit{k}-mer
overlapping them. The vertical dotted line between haplotypes indicates the
location of the causal SNP at this locus. (e) Contingency table of the
phenotypes and haplotypes found in the population at the \textit{G} locus.

% Figure 4 with the results on pubescence density at locus Ps
Figure 4: Results of \textit{k}-mer-based GWAS at the \textit{Ps} locus for
pubescence density.  (a) Genome-wide Manhattan plot showing the location of
significant \textit{k}-mers. The horizontal dotted line indicates the
significance threshold whereas the vertical dotted line indicates the location
of the \textit{Ps} locus.  (b) Zoomed-in Manhattan plot at the \textit{Ps}
locus. The vertical dotted line indicates the location of the most significant
\textit{k}-mer. The shaded gray area shows the location of the causal CNV.  (c)
Estimated copy number (sequencing depth at position divided by accession
average sequencing depth) of accessions with normal and semi-sparse pubescence
at the \textit{Ps} locus. Solid lines represent the median copy number across
73 (semi-sparse pubescence) and 180 accessions (normal pubescence) computed
using a 100-bp sliding window.  Shaded areas outline the first and third
quartiles. The start and end of the CNV region can be clearly inferred from the
divergence between the copy number of accessions with contrasting pubescence.
(d) Histogram showing the location of 1,169 mapped reads containing the most
significant \textit{k}-mer in accessions with semi-sparse pubescence.  Reads
containing this \textit{k}-mer map to either end of the CNV region, suggesting
that they are associated with this variant. Panels (b), (c) and (d) share the
same x-axis.

% Figure 5 with the results on pubescence form at locus Pa1
Figure 5: Results of SNP- and \textit{k}-mer-based GWAS at the \textit{Pa1}
locus for pubescence form. (a) Genome-wide Manhattan plots.  Horizontal dotted
lines indicate the significance threshold whereas vertical dotted lines
indicate the location of the \textit{Pa1} and \textit{Pa2} loci. (b) Zoomed-in
Manhattan plots at the \textit{Pa1} locus.  The vertical dotted lines indicate
the location of the Glyma.12g213900 gene suggested as a candidate for this
locus. The shaded gray rectangles show the regions delimited by the top 1\%
associations.  In the case of SNPs, blue points denote markers used in the
original analysis, whereas red points denote markers that had originally been
pruned but whose \emph{p}-values were computed after signal discovery.  (c)
Gene model of Glyma.12g213900. Rectangles represent exons, colored rectangles
represent coding sequences, and the arrows indicate the direction of
transcription.  (d) Manhattan plots at the location of the Glyma.12g213900
gene.  Vertical lines indicate the location of two non-synonymous SNPs
associated with appressed pubescence, which may represent causal variants at
this locus. Panels (c) and (d) share the same x-axis.

% Figure 6 with the results novel loci for stem termination type
Figure 6: Detection of novel signals for stem termination type using a
\textit{k}-mer-based GWAS analysis contrasting accessions with indeterminate
and semi-determinate stems.  (a) Genome-wide Manhattan plot showing the
location of significant \textit{k}-mers.  The horizontal dotted line indicates
the significance threshold whereas vertical dotted lines indicate the location
of the \textit{Dt1} classical locus as well as three new loci of interest
(\textit{stGm11}, \textit{stGm16}, and \textit{stGm18}).  (b) Pairwise LD
between the 162 significant \textit{k}-mers identified.  \textit{k}-mers are
sorted along the axes according to their putative position on the reference
assembly version 4 of Williams82.  \textit{k}-mers are represented in the same
order along the x- and y-axis. The blue rectangles below the x-axis represent
the -log\textsubscript{10}(\textit{p}) of each \textit{k}-mer.
